Fitting the Undetected: Light-curve Feature Estimation in the Presence of Non-detections

### 2013-03-27 tex/ms.tex
intro paragraphs

...time domain seems simple, because only photometry matters!

...but catalogs make censored point estimates, usually... noise, morphology, CRs, defects...

...catalogs could be better if they did forced photometry.  upper limits worse.
nulls worse.  total obfuscation worst of all

...here we are going to work with nulls; they can still be informative!

@@

Surveys working in real-time tend to analyze each image with an eye to
populating databases quickly.
Unless a particular region of the sky gets special treatment
(such as it is a known place of interest),
discovery of variability happens by comparison of flux values at the database level.
Analysis on such databases of transient/variables thus deals with censored data
where there is often less information available than there should be.
Even then, we might be able to infer what the thresholds were based
on S/N measurements of objects neighboring the source of interest.

>>>>

The measurements of importance for variable-star science are relatively simple.
Unlike the measurements used for astrometry, extragalactic transients, and weak lensing,
where two-dimensional (morphological) information matters,
when variable stars are being observed, all the end user wants is accurate photometry of each source at each epoch.
One could hope that catalog-level information is sufficient:
the detailed image morphology is not required for lightcurve extraction, except when fields are crowded.
However, there are still limitations to working at catalog level:
Catalog entries (nowadays) usually come with noise variance estimates,
but they don't usually give detailed information about the locations or influence of
bad detector pixels, cosmic rays, and flat-field variations;
all these things might be needed to construct a precise photometric noise model or to understand residuals or artifacts.
More importantly, catalogs are usually subject to hard \emph{censoring};
the catalogs (nowadays) only include measurements that are deemed (by the software or catalog makers)
to be \emph{signficant detections}.

This could change.
In principle a catalog could contain a measurement at \emph{every} epoch for any source that is detected at \emph{any} epoch.
This kind of ``forced photometry'' would include many non-significant measurements,
but it would come closer to passing forward all the information in the imaging to users working at catalog level.
We recommend this approach to the catalog makers of the future (especially those making the \textsl{LSST} catalogs).

Less good than the comprehensive forced-photometry approach is to report, when measurements are not significant,
a well-defined ``upper limit'' for the brightness of each source.
This is rarely done, and upper limits are hard to use, but upper limits are an improvement over standard practice.
Standard practice is to \emph{report nothing at all} when a source is not detected significantly.
Indeed, it is often hard, when presented with a multi-epoch catalog,
to determine whether the reason a particular source is missing from the imaging at a particular epoch
is because the source was faint or because the source fell outside the imaging field of view!

In this contribution we look at improving catalog-level time-domain astronomy in the presence of \emph{censored data}.
The problem we consider is that of period and amplitude modeling for large-amplitude variable stars in a data set
in which \emph{no information is reported at all} at the epochs at which the star is not significantly detected.
Furthermore, we will imagine that we also have \emph{no information at all} about the sensitivity of the telescope
at the time of each non-detection; that is, we have no upper limits and we can't reconstruct them from any extant meta-data.
The non-detections are literally database ``null'' values.
This set-up may seem artificial, but it is in fact the situation we find ourselves in for various surveys,
including the ASAS survey, data from which we will model below.
What we will show is that \emph{the ``null'' non-detections are in fact very informative};
amplitude and period fitting will be wrong in general if the non-detections are ignored.

The benefit of using the non-detections is large, but the cost is also large:
Use of these ``null'' data involves non-trivial modeling,
and large numbers of latent parameters to be inferred and marginalized out.

To make this all concrete, consider the photometric data from the \textsl{ASAS} survey
depicted in \figurename~\ref{fig:example} for one Mira variable.
The photometric detections are shown and so are the epochs at which there are nulls (or non-detections).
The standard

@@
